---
title: "STAT40830 - Homework 1"
author: "Padraig Mac Suibhne - 23201326 "
date: today
date-format: long
editor: visual
toc: true
format: 
  html:
   embed-resources: true
  pdf: 
    geometry:
    - top=18mm
    - left=15mm
    - bottom=25mm
    - heightrounded
    fontsize: 11pt
    linestretch: 1.5
execute: 
  error: true
echo: false
---

\newpage

## 1. Introduction

In this exercise, we will plot the \textbf{\textcolor{purple}{Nile}} data set that provides measurements of the annual flow of the river Nile at Aswan (formerly Assuan), 1871–1970, in $10^8 m^3$ “with apparent changepoint near 1898” (Cobb(1978), Table 1, p.249).

This is a new line to demonstrate a committed change.

```{r}
#| label: load-libraries-data-prep
#| message: false
library(knitr)
library(ggplot2)

Nile_summary_array <- as.array(summary(Nile))
dimnames(Nile_summary_array) <- list(c("Minimum", "First Quartile", "Median", "Mean", "Third Quartile", "Maximum"))
# Convert array to a data frame
Nile_summary_df <- data.frame(Statistic = names(Nile_summary_array), Annual_Flow = as.vector(Nile_summary_array))

```

## 2. Data Summary

```{r}
#| label: tbl-nile-data-summary
#| tbl-cap: "Summary of Nile Data Set"
#| tbl-colwidths: [20,15]
#| layout-ncol: 1


# Display the table with kable
kable(Nile_summary_df)
```

The **Nile** data set provides **100** measurements, the summary of which is provided in @tbl-nile-data-summary.

@fig-nile-time-series-plot outlines the annual flow rate for the timespan of the data set.

```{r}
#| label: fig-nile-time-series-plot
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Annual Flow of River Nile 1871-1970"


# Convert Nile dataset into a dataframe with a numeric year
nile_df <- data.frame(Year = as.numeric(time(Nile)), Flow = as.numeric(Nile))

# Plot with ggplot2 without the warning
ggplot(nile_df, aes(x = Year, y = Flow)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  labs(title = "Annual Flow Nile River",
       x = "Year",
       y = "Flow (10e8 cubic meters)") +
  theme_minimal()
```

## 3. Data Review

Based on the reference from Cobb (1978), there may be a change point in 1898. We will now investigate this in this section.

```{r}
#| label: investigate-1898
#| message: false


# Pre-1898 data
pre_1898_data <- Nile[time(Nile) <= 1898]
pre_summary <- summary(pre_1898_data)

# Post-1898 data
post_1898_data <- Nile[time(Nile) > 1898]
post_summary <- summary(post_1898_data)


```

```{r}
#| label: tbl-nile-pre-post
#| tbl-cap: "Annual Flow Pre and Post 1898"
#| tbl-colwidths: [20,15,15]
#| layout-ncol: 1

# Create a data frame for display
summary_table <- data.frame(
  Statistic = names(Nile_summary_array),
  Pre_1898 = round(as.numeric(pre_summary),2),
  Post_1898 = round(as.numeric(post_summary),2)
)

# Display the table using kable
kable(summary_table)
```

From @tbl-nile-pre-post , we can make the following conclusions:

-   the annual flow rate post 1898 has seen a significant reduction on the data pre 1898.
-   We see the median flow post 1898 is **`r paste(round(post_summary["Median"]/pre_summary["Median"]*100,1),"%", sep="")`** of the flow pre-1898.
-   the mean annual flow rate has also reduced by *`r paste(round(100-post_summary["Mean"]/pre_summary["Mean"]*100,1),"%", sep="")`*, which is consistent with the \textcolor{red}{reduction in the median value}.
-   \textcolor{blue}{you will notice that the reference to Table 3 is incorrect. In }\textcolor{green}{HTML format }\textcolor{blue}{ this table is designated as Table 2, but unfortunately in }\textcolor{green}{PDF format }\textcolor{blue}{ the reference is Table 3. I tried unsuccessfully to fix this.}

## 4. References

Cobb, G. W. (1978). The problem of the Nile: conditional solution to a change-point problem. Biometrika 65, 243–51.
